### 2016-05-12 text/AppendixA.tex
Context-aware systems edits (Chapter 3)

The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
maximal number of past values required by a data-flow computation. The discussion includes the 
following example:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
In order to give typing for the above example, we first need to extend the language with 
conditionals. The typing rule for the $\kvd{if}$ construct is:
%
\begin{equation*}
\tyrule{if}
  { \coctx{\Gamma}{\cclrd{m}} \vdash e : \ident{bool} &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_1 : \tau &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau }
  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m},\cclrd{n})} \vdash \kvd{if}~e~\kvd{then}~e_1~\kvd{else}~e_2 : \tau_1 }
\end{equation*}
%
Given a condition that requires $\cclrd{m}$ past and two (alternative) bodies that each require
$\cclrd{n}$ past values, the composed expression requires at most the maximum of the two, \ie~the
same context is passed to both the condition and the body using point-wise composition. As we will
see, the fact that both branches require the same number of past values means that we need to use
the sub-coeffecting rule.

Assuming $\Gamma_0 = \ident{tick}:\ident{int}, \ident{counter}:\ident{int}$, the 
Figure~\ref{fig:appendixa-dfl} shows the typing derivation for the example. In the first part,
we derive the types for the sub-expressions of the application and the conditional. Note that 
to obtain compatible function types, we use sub-coeffecting \emph{before} abstraction in the
typing of $(\lambda x.x)$ -- the type $\ident{int} \xrightarrow{\cclrd{1}} \ident{int}$ is  a
valid overapproximation. Finally, the typing of the application yields the requirement of two 
past values, calculated as $\textnormal{max}(\cclrd{1}, \cclrd{1}+\cclrd{1})$.

>>>>

The Section~\ref{sec:applications-flat-dataflow} presents a coeffect type system that tracks the 
maximal number of past values required by a data-flow computation. The discussion includes the 
following example:
%
\begin{equation*}
\begin{array}{l}
(\,   \kvd{if}~~(\kvd{prev}~\ident{tick})=0\\[-0.25em]
\,\,\, \kvd{then}~(\lambda x \rightarrow \kvd{prev}~x)\\[-0.25em]
\,\,\, \kvd{else}~(\lambda x \rightarrow x) \,)\qquad(\kvd{prev}~\ident{counter})
\end{array}
\end{equation*}
%
In order to give typing for the above example, we first need to extend the language with 
conditionals. The typing rule for the $\kvd{if}$ construct is:
%
\begin{equation*}
\tyrule{if}
  { \coctx{\Gamma}{\cclrd{m}} \vdash e : \ident{bool} &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_1 : \tau &
    \coctx{\Gamma}{\cclrd{n}} \vdash e_2 : \tau }
  {\coctx{\Gamma}{\textnormal{max}(\cclrd{m},\cclrd{n})} \vdash \kvd{if}~e~\kvd{then}~e_1~\kvd{else}~e_2 : \tau_1 }
\end{equation*}
%
Given a condition that requires $\cclrd{m}$ past and two (alternative) bodies that each require
$\cclrd{n}$ past values, the composed expression requires at most the maximum of the two, \ie~the
same context is passed to both the condition and the body using point-wise composition. As we will
see, the fact that both branches require the same number of past values means that we need to use
the subcoeffecting rule.

Assuming $\Gamma_0 = \ident{tick}:\ident{int}, \ident{counter}:\ident{int}$, the 
Figure~\ref{fig:appendixa-dfl} shows the typing derivation for the example. In the first part,
we derive the types for the sub-expressions of the application and the conditional. Note that 
to obtain compatible function types, we use subcoeffecting \emph{before} abstraction in the
typing of $(\lambda x.x)$ -- the type $\ident{int} \xrightarrow{\cclrd{1}} \ident{int}$ is  a
valid overapproximation. Finally, the typing of the application yields the requirement of two 
past values, calculated as $\textnormal{max}(\cclrd{1}, \cclrd{1}+\cclrd{1})$.

### 2016-05-12 text/AppendixB.tex
Context-aware systems edits (Chapter 3)

\paragraph{(var)} Given the following derivation using (\emph{var}):
\[
\inference
  { }
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit} \vdash y : \tau }
\]
There are two cases depending on whether $y$ is the variable $x$ or not:
\begin{itemize}
\item If $y=x$ then also $\tau = \tau_s$ and thus $\subst{y}{x}{e_s} = e_s$. Using the assumption,
implicit weakening and the fact that $\cunit$ is a unit of $\cseq$:
\[
\inference
 {\inference 
  {\coctx{\Gamma}{\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau_s }
  {\coctx{\Gamma_1, \Gamma, \Gamma_2}{\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau } }
 {\coctx{\Gamma_1, \Gamma, \Gamma_2}{\cunit\,\cseq\,\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau } 
\]
\item If $y\neq x$ then $\subst{y}{x}{e_s} = y$. Using the fact that $\cunit$ is the bottom element
and sub-coeffecting:
\[
\inference
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit} \vdash y : \tau }
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit\,\cseq\,\cclrd{S}} \vdash y : \tau }
\]
\end{itemize}

@@

\paragraph{(sub)} As in the (\emph{contr}) case, in the (\emph{sub}) case we distinguish two situations.
If the sub-coeffecting is applied to variable that is \emph{not} being substituted for, then the case is
easy (sub-coeffecting does not interact with substitution), so we only consider the case when $x$ is one
of the variables being substituted for:
\[
\inference
  {{\begin{array}{l}
   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
     \\[-0.35em] \qquad
     \coctx{x\!:\!\tau_1}{\alift{\cclrd{s'}}}
     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
   \vdash e : \tau
   \end{array}}}
  {{\begin{array}{l}
   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
     \\[-0.35em] \qquad
     \coctx{x\!:\!\tau_1}{\alift{\cclrd{s}}}
     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
   \vdash e : \tau
   \end{array}}}
\quad(\cclrd{s'}\;\cleq\;\cclrd{s})   
\]
From the induction hypothesis, we have the following:
\[
  {{\begin{array}{l}
   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\,
     \\[-0.35em] \qquad
     \coctx{\Gamma_{Ti}}{\cclrd{s'}\aseq\,\aclrd{\mathbf{T_k}} }
     \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
   \vdash e : \tau
   \end{array}}}
\]
To complete the case, we need to apply (\emph{sub}) repeatedly on each of the variables in 
$\Gamma_{Ti}$. For $i^{\textit{th}}$ variable $x_i$, the coeffect annotation is $\cclrd{s'}\cseq\,\cclrd{T_i}$.
Using the fact that combining coeffects with $\cseq$ preserves the ordering, we get that 
$(\cclrd{s'}\cseq\,\cclrd{T_i}) \cleq (\cclrd{s}\,\cseq\,\cclrd{T_i})$ and so the conditions
of (\emph{sub}) are satisfied.

\paragraph{(weak)} We again need to consider whether the removed variable is 
one of the variables that are being substituted for. If this is not the case,
the proof is easy, so we only look at the other case:
\[
\inference
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} }~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
   \vdash e : \tau}
  {\coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,|\,\coctx{x\!:\!\tau_1}{\czero}\,]
   \vdash e : \tau} 
\]
Now, we use the induction hypothesis, apply the (\emph{weak}) rule and use properties of the
structural coeffect algebra:
\[
\inference
  {{ \begin{array}{l}
    \coctx{\Gamma}{ \aclrd{\textbf{r}} }~
    [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
     \\[-0.35em] \qquad
     \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}\,]
   \vdash e~[\,\ldots\,] : \tau 
   \end{array} }}
{\tyrule{sub}
 {{\begin{array}{l}
   \coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
    \\[-0.35em] \qquad
    \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}
      \,|\,\coctx{x\!:\!\tau_1}{\czero}\,]
   \vdash e~[\,\ldots\,] : \tau
   \end{array} }} 
 {{\begin{array}{l}
   \coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
    \\[-0.35em] \qquad
    \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}
      \,|\,\coctx{x\!:\!\tau_1}{\czero \aseq\,\aclrd{\textbf{T}_k} }\,]
   \vdash e~[\,\ldots\,] : \tau
   \end{array} }}  }
\]
The derivation first applies the standard (\emph{weak}) rule and then uses sub-coeffecting
rule and the property $\czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})$ to obtain conclusion of the required form.

>>>>

\paragraph{(var)} Given the following derivation using (\emph{var}):
\[
\inference
  { }
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit} \vdash y : \tau }
\]
There are two cases depending on whether $y$ is the variable $x$ or not:
\begin{itemize}
\item If $y=x$ then also $\tau = \tau_s$ and thus $\subst{y}{x}{e_s} = e_s$. Using the assumption,
implicit weakening and the fact that $\cunit$ is a unit of $\cseq$:
\[
\inference
 {\inference 
  {\coctx{\Gamma}{\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau_s }
  {\coctx{\Gamma_1, \Gamma, \Gamma_2}{\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau } }
 {\coctx{\Gamma_1, \Gamma, \Gamma_2}{\cunit\,\cseq\,\cclrd{S}} \vdash \subst{y}{x}{e_s} : \tau } 
\]
\item If $y\neq x$ then $\subst{y}{x}{e_s} = y$. Using the fact that $\cunit$ is the bottom element
and subcoeffecting:
\[
\inference
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit} \vdash y : \tau }
  {\coctx{\Gamma_1, y:\tau, \Gamma_2}{\cunit\,\cseq\,\cclrd{S}} \vdash y : \tau }
\]
\end{itemize}

@@

\paragraph{(sub)} As in the (\emph{contr}) case, in the (\emph{sub}) case we distinguish two situations.
If the subcoeffecting is applied to variable that is \emph{not} being substituted for, then the case is
easy (subcoeffecting does not interact with substitution), so we only consider the case when $x$ is one
of the variables being substituted for:
\[
\inference
  {{\begin{array}{l}
   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
     \\[-0.35em] \qquad
     \coctx{x\!:\!\tau_1}{\alift{\cclrd{s'}}}
     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
   \vdash e : \tau
   \end{array}}}
  {{\begin{array}{l}
   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
     \\[-0.35em] \qquad
     \coctx{x\!:\!\tau_1}{\alift{\cclrd{s}}}
     \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
   \vdash e : \tau
   \end{array}}}
\quad(\cclrd{s'}\;\cleq\;\cclrd{s})   
\]
From the induction hypothesis, we have the following:
\[
  {{\begin{array}{l}
   \coctx{\Gamma_1,-,\Gamma_2}{\aclrd{\textbf{r}} \atimes \alift{-} \atimes \aclrd{\textbf{q}}} ~
   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\,
     \\[-0.35em] \qquad
     \coctx{\Gamma_{Ti}}{\cclrd{s'}\aseq\,\aclrd{\mathbf{T_k}} }
     \,|\, \ldots \,|\, \coctx{\Gamma_{Tk}}{\cclrd{s_k}\aseq\,\aclrd{\textbf{T}_k}}\,]
   \vdash e : \tau
   \end{array}}}
\]
To complete the case, we need to apply (\emph{sub}) repeatedly on each of the variables in 
$\Gamma_{Ti}$. For $i^{\textit{th}}$ variable $x_i$, the coeffect annotation is $\cclrd{s'}\cseq\,\cclrd{T_i}$.
Using the fact that combining coeffects with $\cseq$ preserves the ordering, we get that 
$(\cclrd{s'}\cseq\,\cclrd{T_i}) \cleq (\cclrd{s}\,\cseq\,\cclrd{T_i})$ and so the conditions
of (\emph{sub}) are satisfied.

\paragraph{(weak)} We again need to consider whether the removed variable is 
one of the variables that are being substituted for. If this is not the case,
the proof is easy, so we only look at the other case:
\[
\inference
  {\coctx{\Gamma}{ \aclrd{\textbf{r}} }~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,]
   \vdash e : \tau}
  {\coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
   [\,\coctx{x_{T1}\!:\!\tau_{T1}}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\, \coctx{x_{Tk}\!:\!\tau_{Tk}}{\alift{\cclrd{s_k}}}\,|\,\coctx{x\!:\!\tau_1}{\czero}\,]
   \vdash e : \tau} 
\]
Now, we use the induction hypothesis, apply the (\emph{weak}) rule and use properties of the
structural coeffect algebra:
\[
\inference
  {{ \begin{array}{l}
    \coctx{\Gamma}{ \aclrd{\textbf{r}} }~
    [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
     \\[-0.35em] \qquad
     \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}\,]
   \vdash e~[\,\ldots\,] : \tau 
   \end{array} }}
{\tyrule{sub}
 {{\begin{array}{l}
   \coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
    \\[-0.35em] \qquad
    \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}
      \,|\,\coctx{x\!:\!\tau_1}{\czero}\,]
   \vdash e~[\,\ldots\,] : \tau
   \end{array} }} 
 {{\begin{array}{l}
   \coctx{\Gamma, -}{\aclrd{\textbf{r}} \atimes \alift{ - }}~
   [\,\coctx{\Gamma_{T1}}{\cclrd{s_1}\aseq\,\aclrd{\textbf{T}_1}} \,|\, \ldots \,|\, 
    \\[-0.35em] \qquad
    \coctx{\Gamma_{Tk-1}}{\cclrd{s_{k-1}}\aseq\,\aclrd{\textbf{T}_{k-1}}}
      \,|\,\coctx{x\!:\!\tau_1}{\czero \aseq\,\aclrd{\textbf{T}_k} }\,]
   \vdash e~[\,\ldots\,] : \tau
   \end{array} }}  }
\]
The derivation first applies the standard (\emph{weak}) rule and then uses subcoeffecting
rule and the property $\czero\;\cleq\;(\czero\,\cseq\,\cclrd{r})$ to obtain conclusion of the required form.
